---
title: "Regularization & Feature Selection"
author: "Tuba Aksoy"
output: pdf_document
date: "2023-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Data Analysis. For this problem, you may use one of two data sets: 
(i) the Boston House Prices dataset available in scikit-learn (load boston) or (ii) the Baseball dataset available from the ISLR
R package (install(ISLR) & data(Hitters)). The response for these datasets is MEDV, the median
value of home prices, or Salary respectively. Note that if the response is highly skewed, then you may
want to take the logarithm of the response. Additionally, if there are any missing values, you may want
to remove these rows for this problem. Please document how you processed this data for analysis.

### (a) What are the most important features?

### (i) Compare and contrast the top features as determined by:

1- Statistical significance in Linear Regression.
2- Best Subsets.
3- Step-wise approaches (and/or Recursive Feature Elimination).
4- Lasso.
5- Elastic Net.

### (ii) Fit and visualize regularization paths for the following methods:

1- Lasso
2- Elastic Net (at two separate alpha's).
3- Ridge.

### (iii) Reflect on these results.

Are the top features different for each method? Why or why not? If there are tuning parameters, how did you determine these? Do different tuning
parameters yield different important features? Are there any features consistently selected
by all methods? What are the most important features and how did you determine this?
Explain and expand on your responses.



### (b) Which linear method is best for prediction? 

For this part of the problem, you will need to randomly split the data into 60% training, 20% validation and 20% test sets. If methods have any
tuning parameters, you should fit models on the training data, choose the tuning parameters by
minimizing the prediction error on the validation set, and then report the final prediction error
on the test set. Repeat this procedure 10 times and average the results.

### (i) Compare the average prediction MSE on the test set for the following methods:

1-Least Squares.
2- Ridge Regression.
3- Best Subsets.
4- Step-wise approaches (and/or Recursive Feature Elimination).
5- Lasso.
6- Elastic Net.

### (ii) Visualize the results of your comparisons. 

You can choose to only show the results with the best tuning parameter values.

### (iii) Reflection. 

Which types of methods give the best prediction error? Why do these methods
perform well? Do any methods seem to overfit to the training set? If so, why? Do all the
methods that give similar predictions choose the same subset of variables? Which is the
overall best method for prediction on this dataset? Explain and expand on your responses.

### 2. Empirical Demonstrations of Regression Properties. 

For these problems, you can use the data mentioned above or you may need to design your own linear regression simulation.

### (a) Empirically demonstrate that fitting linear regression with an intercept term is equivalent to (i)fitting linear regression when centering Y and centering the columns of X, and (ii) fitting linear regression when adding a column of ones to X.


### (b) Empirically demonstrate that the least squares solution has zero training error when p > n.

### (c) Empirically demonstrate the MSE Existence Theorem. 

Recall that the MSE Existence Theorem states that there exists a value of lamda for which MSE(Xbetahat^Ridge(lamda) < MSE(XBetahat^LS).


\newpage

```{r part0}

library(ISLR2)
library(glmnet)
library(leaps)
library(MASS)
library(stats)
library(tidyverse)
library(caret)
library(leaps)
library(dplyr)
library(purrr)
library(ggplot2)
library(caret)
data("Hitters")  
### CLEAN / LOG TRANSFORM /ORGANIZE
cleaned_hitters <- na.omit(Hitters)
cleaned_hitters$Log_Salary <- log(cleaned_hitters$Salary) 
numerical_data_1 <- cleaned_hitters[, c("AtBat", "Hits", "HmRun", "Runs", "RBI", "Walks", "Years", "CAtBat")]
numerical_data_2 <- cleaned_hitters[, c("CHits", "CHmRun", "CRuns", "CRBI", "CWalks", "PutOuts", "Assists", "Errors")]
categorical_data <- cleaned_hitters[, c("League", "Division", "NewLeague")]
dumm_data <- dummyVars(~., data = categorical_data)
encoded_data <- data.frame(predict(dumm_data, newdata = categorical_data))
Log_Salary<- cleaned_hitters[, "Log_Salary"]
Salary<- cleaned_hitters[, "Salary"]
organized_data <- cbind(Log_Salary, numerical_data_1, numerical_data_2, encoded_data)
organized_data1 <- cbind(Salary, numerical_data_1, numerical_data_2, encoded_data)
N <- nrow(organized_data)
### PAIRS
numerical_p_1 <- cleaned_hitters[, c("AtBat", "Hits", "HmRun", "Runs", "RBI", "Walks", "Years", "CAtBat", "Salary")]
numerical_p_2 <- cleaned_hitters[, c("CHits", "CHmRun", "CRuns", "CRBI", "CWalks", "PutOuts", "Assists", "Errors", "Salary")]
categorical_p <- cleaned_hitters[, c("League", "Division", "NewLeague", "Salary")]
###SPLIT AND SCALE THE DATA###
N <- nrow(organized_data)
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)
DATA_SCALED<- rbind(data_train_scaled,data_valid_scaled,data_test_scaled)
y_centered = as.numeric(DATA_SCALED[, 1])
X_standardized = as.matrix(DATA_SCALED[, -1])
```



\newpage
```{r part1, fig.width=4, fig.height=3}

hist(cleaned_hitters$Salary, main="Salary Histogram", xlab="Salary") # Histogram of the original Salary variable
hist(cleaned_hitters$Log_Salary, main="Log_Salary Histogram", xlab="Log(Salary)") # Histogram of the Log_Salary variable

```
\newpage



```{r part2}
# Create scatterplot matrices for both groups
pairs(numerical_p_1)
pairs(numerical_p_2)
pairs(categorical_p)

```
\newpage



```{r part3}
### LEAST SQUARES- LINEAR REGRESSION::
fitc = lm(y_centered~X_standardized)
#most important features
significance_level <- 0.05
summary_lm <- summary(fitc)
stripped_variable_names <- sub("^X_standardized", "", rownames(summary_lm$coefficients[summary_lm$coefficients[, 4] < significance_level, , drop = FALSE]))

```


```{r part4}
### BEST SUBSETS
fitbsub = regsubsets(x = X_standardized, y = y_centered)
res.sum<-summary(fitbsub)
# WHICH FEATURES
ADJ = which.max(res.sum$adjr2)
CP = which.min(res.sum$cp)
BIC = which.min(res.sum$bic)
get_features <- function(id, object, outcome){
  models <- summary(object)$which[id,-1]
    predictors <- names(which(models == TRUE))
  predictors <- paste(predictors, collapse = "+")
}

```

```{r part5}
### FORWARD STEPWISE
# (via CP)
fit0 = lm(Log_Salary ~ 1, data = DATA_SCALED)
formula <- as.formula(paste("Log_Salary ~", paste(setdiff(colnames(data_train_scaled), "Log_Salary"), collapse = " + ")))
fitf = stepAIC(fit0, scope = formula, direction = "forward", data = DATA_SCALED, k = log(nrow(DATA_SCALED)))

### BACKWARD STEPWISE
# (via AIC)
fit0 = lm(Log_Salary ~ ., data = DATA_SCALED)
fitb = stepAIC(fit0, direction = "backward", data = DATA_SCALED, k = log(nrow(DATA_SCALED)))


Forward_features <- names(coef(fitf))[!coef(fitf) == 0][-1]
Backward_features <- names(coef(fitb))[!coef(fitb) == 0][-1]


```
\newpage
```{r part6}

### LASSO
fit0 = lm(y_centered ~ X_standardized - 1)


lam = 0.2 # alpha = 1 stands for Lasso regression
fitl_02 = glmnet(x = X_standardized, y = y_centered, family = "gaussian",
              lambda = lam, alpha = 1, intercept = FALSE)


coefs02 = cbind(fit0$coef, fitl_02$beta)
colnames(coefs02) = c("Without L1", "With L1")
LASSO_features2 <- rownames(as.matrix(round(coefs02, 10)))[as.matrix(round(coefs02, 10))[, 2] != 0]




lam = 0.3 # alpha = 1 stands for Lasso regression
fitl_03 = glmnet(x = X_standardized, y = y_centered, family = "gaussian",
                           lambda = lam, alpha = 1, intercept = FALSE)


coefs03 = cbind(fit0$coef, fitl_03$beta)
colnames(coefs03) = c("Without L1", "With L1")
LASSO_features <- rownames(as.matrix(round(coefs03, 10)))[as.matrix(round(coefs03, 10))[, 2] != 0]

```
\newpage
```{r part7}
###Elastic Net
fit0 = lm(y_centered ~ X_standardized - 1)

lam1 = 0.1
fitl1 = glmnet(x = X_standardized, y = y_centered, family = "gaussian",
               lambda = lam1, alpha = 0.5, intercept = FALSE)


lam2 = 0.9
fitl2 = glmnet(x =X_standardized , y = y_centered, family = "gaussian",
               lambda = lam2, alpha = 0.5, intercept = FALSE)


# Extract important features
coefs_EL = cbind(fit0$coef, fitl1$beta, fitl2$beta)
colnames(coefs_EL) = c("lambda = 0", "lambda = 0.1", "lambda = 0.9")
coefficients_matrix3 <- as.matrix(round(coefs_EL, 10))
important_features3 <- rownames(coefficients_matrix3)[coefficients_matrix3[, 2] != 0]
important_features4 <- rownames(coefficients_matrix3)[coefficients_matrix3[, 3] != 0]

coefs_EL

```

\newpage




### (ii) Fit and visualize regularization paths for the following methods:

### 1- Lasso

```{r part8}
# Lasso paths

# Lasso paths
fitl = glmnet(x = X_standardized, y = y_centered, family = "gaussian", alpha = 1)
plot(fitl)
legend("topleft", legend = colnames(X_standardized), col = 1:23, lty = rep(1, 8), cex = 0.8)
title("LASSO PATHS")

```
\newpage

### 2- Elastic Net (at two separate alpha's).

```{r part9}
alphas = c(0.3, 0.7)

# Loop through each alpha value, fit the model, and plot the paths
par(mfrow = c(1, 1))
for (alpha in alphas){
    glm_fit = glmnet(x = X_standardized, y = y_centered, alpha = alpha)
    plot(glm_fit, xvar = "lambda", main = paste("ELASTIC NET", "alpha =", alpha))
    legend("topright", legend = colnames(X_standardized), col = 1:23, lty = 1, cex = .8)
}


```
\newpage

### 3- Ridge.

```{r part10}
fit_glm = glmnet(X_standardized, y_centered,alpha=0,standardize=FALSE,intercept=FALSE)
plot(fit_glm,xvar="lambda")
legend("topright",legend=names(organized_data)[2:17],col=1:23,lty=rep(1,17))

```
\newpage



### (a) What are the most important features?

### (i) Compare and contrast the top features as determined by:

### 1- Statistical significance in Linear Regression.

### 2- Best Subsets.

### 3- Step-wise approaches (and/or Recursive Feature Elimination).

### 4- Lasso.

### 5- Elastic Net.


### (iii) Reflect on these results.

Are the top features different for each method? Why or why not? If there are tuning parameters, 
how did you determine these? Do different tuning
parameters yield different important features? Are there any features consistently selected
by all methods? What are the most important features and how did you determine this?
Explain and expand on your responses.

```{r part77}


## ANSWER:


cat("Here are the MOST IMPORTANT FEATURES for different Methods/ parameters:", "\n",
"Statistical significance in LINEAR REGRESSION:", stripped_variable_names, "\n",
"BEST SUBSET with Adj.R2:", get_features(ADJ, fitbsub, "Log_Salary"), "\n",
"BEST SUBSET with CP:", get_features(CP, fitbsub, "Log_Salary"), "\n",
"BEST SUBSET with BIC:", get_features(BIC, fitbsub, "Log_Salary"), "\n",
"FORWARD:", paste(Forward_features, collapse = " + "), "\n",
"BACKWARD:", paste(Backward_features, collapse = " + "), "\n",
"LASSO (LAMDA=0.2):" ,LASSO_features2,  "\n",
"LASSO (LAMDA=0.3):" ,LASSO_features,  "\n",
"ELASTIC NET (LAMDA=0.1, alpha=0.5):" ,important_features3,  "\n",
"ELASTIC NET (LAMDA=0.9, alpha=0.5):" ,important_features4,  "\n",
"The top features are different for each method and for different tuning parameters.","\n",
"That is because methods use different mechanisms and criteria for feature selection ","\n",
"and regularization.","\n",
"Selection of the model, number of features, and tuning parameters depend on the goal of the study", "\n",
"and requires some background knowledge.","\n",
"For this homework, I picked the parameters to compare for different distances of parameters.", "\n", 
"For example in elastic net, I put alpha=0.5, bacecause I wanted to be neither close ", "\n", 
"to Ridge nor Lasso, and compared two LAMDA values to see how much the increase affect the changes" ,"\n",
"Hit, Walks, and Years are 3 features that are most frequenltly selected.", "\n",
"To determine the real best features I would get more idea of the context of the data/ background ", "\n",
"and determine either a limit to the number of features, or focus on models that enhances the", "\n",
"essential features. Choosing the model also depend on the success of the method will in ", "\n",
"predicting the target variable. ", "\n")


```

\newpage

### (b) Which linear method is best for prediction? 

For this part of the problem, you will need to randomly split the data into 60% training, 
20% validation and 20% test sets. If methods have any tuning parameters, you should fit models on the 
training data, choose the tuning parameters by minimizing the prediction error on the validation set, 
and then report the final prediction error on the test set. Repeat this procedure 10 times and average the results.

### (i) Compare the average prediction MSE on the test set for the following methods:
### 1-Least Squares.
### 2- Ridge Regression.
### 3- Best Subsets.
### 4- Step-wise approaches (and/or Recursive Feature Elimination).
### 5- Lasso.
### 6- Elastic Net.

```{r part12}
### 1-Least Squares.
# Initialize variables to store the average MSE across iterations
LS_test_MSE_mean <- 0
num_iterations <- 10


for (ii in 1:num_iterations) {
#Split 
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)


#TRAIN#
LS_model <- lm(y_train_centered ~ ., data = data_train_scaled)

#VALID#
LS_valid_predictions <- predict(LS_model, newdata = data_valid_scaled)
LS_valid_mse <- mean((LS_valid_predictions - y_valid_centered)^2)
  
#TEST#
LS_test_predictions <- predict(LS_model, newdata = data_test_scaled)
LS_test_mse <- mean((LS_test_predictions - y_test_centered)^2)
LS_test_MSE_mean <- LS_test_MSE_mean + LS_test_mse
}

# Calculate the average test MSE across all iterations for Least Squares
MEAN_MSE_LS <- LS_test_MSE_mean / num_iterations

```

```{r part13}
### 2- Ridge Regression.
num_iterations <- 10
RIDGE_BEST_alpha_values <- numeric(num_iterations)
RIDGE_BEST_mse_values <- numeric(num_iterations)
RIDGE_test_MSE_mean <- 0

for (ii in 1:num_iterations) {

#Split 
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- (scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- (scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- (scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)

#TRAIN#
alpha_values <- seq(0.1, 1, by = 0.1)
RIDGE_model_list <- list()  # To store models for different alpha values

for (alpha in alpha_values) {
  RIDGE_model <- glmnet(x = X_train_scaled, y = y_train_centered, alpha = alpha)
  RIDGE_model_list[[as.character(alpha)]] <- RIDGE_model
}

#VALID#
RIDGE_mse_values <- numeric(length(alpha_values))

for (i in seq_along(alpha_values)) {
  RIDGE_valid_predictions <- predict(RIDGE_model_list[[as.character(alpha_values[i])]], newx = X_valid_scaled    )
  RIDGE_mse_values[i] <- mean((RIDGE_valid_predictions - y_valid_centered)^2)

}

RIDGE_best_alpha <- alpha_values[which.min(RIDGE_mse_values)]
Ridge_best_model <- RIDGE_model_list[[as.character(RIDGE_best_alpha)]]
RIDGE_best_mse<- min(RIDGE_mse_values)

# Store the results
RIDGE_BEST_alpha_values[ii] <- RIDGE_best_alpha
RIDGE_BEST_mse_values[ii] <- RIDGE_best_mse

#TEST#
RIDGE_test_predictions <- predict(Ridge_best_model, newx = X_test_scaled)
RIDGE_test_mse <- mean((RIDGE_test_predictions - y_test_centered)^2)
# Add the test MSE for this iteration to the running total
RIDGE_test_MSE_mean <- RIDGE_test_MSE_mean + RIDGE_test_mse
}

# Calculate the average test MSE across all iterations for Least Squares
MEAN_MSE_RIDGE <- RIDGE_test_MSE_mean / num_iterations
table_RIDGE <- data.frame(Iteration = 1:num_iterations, Best_Validation_Alpha = RIDGE_BEST_alpha_values, Best_Validation_MSE = RIDGE_BEST_mse_values)
#print(table_RIDGE)

```

```{r part14}
### 3- Best Subsets.
model1_count <- 0
model2_count <- 0
model3_count <- 0
mse_values <- numeric(10)  # Assuming you want to run the program 10 times
BESTSUM_test_MSE_mean <- 0

for (ii in 1:10) { 

#Split 
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)


#TRAIN#
BESTSUB_model = regsubsets(x = X_train_scaled, y = y_train_centered)
res.sum = summary(BESTSUB_model)
ADJ = which.max(res.sum$adjr2)
CP = which.min(res.sum$cp)
BIC = which.min(res.sum$bic)

get_model_formula <- function(id, object, outcome){
  models <- summary(object)$which[id,-1]
    predictors <- names(which(models == TRUE))
  predictors <- paste(predictors, collapse = "+")
  as.formula(paste0(outcome, "~", predictors))
}

formula_Adj<- get_model_formula(ADJ, BESTSUB_model, "Log_Salary")
formula_CP<- get_model_formula(CP, BESTSUB_model, "Log_Salary")
formula_BIC<- get_model_formula(BIC, BESTSUB_model, "Log_Salary")

#VALIDATION 
model1 <- lm(formula_Adj, data = data_valid_scaled)
model2<- lm(formula_CP, data=data_valid_scaled)
model3<- lm(formula_BIC, data=data_valid_scaled)
BESTSUB_valid_predictions_adj <- predict(model1, newx = X_valid_scaled)
BESTSUB_valid_predictions_cp <- predict(model2, newx = X_valid_scaled)
BESTSUB_valid_predictions_bic <- predict(model3 , newx = X_valid_scaled)
mse_model1 <- mean((BESTSUB_valid_predictions_adj - y_valid_centered)^2)
mse_model2 <- mean((BESTSUB_valid_predictions_cp - y_valid_centered)^2)
mse_model3 <- mean((BESTSUB_valid_predictions_bic - y_valid_centered)^2)
 
# Determine the best model based on the lowest MSE
if (mse_model1 < mse_model2 && mse_model1 < mse_model3) {
  best_model <- model1
  best_model_name <- "Model 1"
  model1_count <- model1_count + 1
} else if (mse_model2 < mse_model1 && mse_model2 < mse_model3) {
  best_model <- model2
  best_model_name <- "Model 2"
  model1_count <- model1_count + 1
} else {
  best_model <- model3
  best_model_name <- "Model 3"
  model1_count <- model1_count + 1
}


#TEST
BESTSUB_test_predictions <- predict(best_model, newdata = data_test_scaled)
BESTSUB_test_mse <- mean((BESTSUB_test_predictions - data_test_scaled$Log_Salary) ^ 2)
BESTSUM_test_MSE_mean <- BESTSUM_test_MSE_mean + BESTSUB_test_mse
}

MEAN_MSE_BESTSUBSET <- BESTSUM_test_MSE_mean / num_iterations
model_names <- c("Model ADJ", "Model CP", "Model BIC")
selection_counts <- c(model1_count, model2_count, model3_count)
table_BESTSUBSET <- data.frame(Model = model_names, Selection_Count = selection_counts)
#print(table_BESTSUBSET)

```


```{r part15,results="hide"}
### 4- Step-wise approaches (and/or Recursive Feature Elimination).
##FORWARD
FORWARD_test_MSE_mean<- 0
aic_count_forward <- 0
bic_count_forward <- 0


for (ii in 1:num_iterations) {
#Split 
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)

#TRAINING
#AIC
fit0 = lm(Log_Salary ~ 1, data = data_train_scaled)
column_names <- colnames(data_train_scaled)
predictor_names <- setdiff(column_names, "Log_Salary")
formula <- as.formula(paste("Log_Salary ~", paste(predictor_names, collapse = " + ")))
fitf = stepAIC(fit0, scope =formula ,
               direction = "forward", data = data_train_scaled, k = log(nrow(data_train_scaled)),  verbose = FALSE)
selected_variables <- names(coef(fitf))[!coef(fitf) == 0]
ForwardVariables <- selected_variables[-1]
formula_1 <- as.formula(paste("Log_Salary ~", paste(ForwardVariables, collapse = " + ")))

 #BIC 
fitf_bic = stepAIC(fit0, scope =formula ,
               direction = "forward", data = data_train_scaled, k = log(nrow(data_train_scaled)) * (log(nrow(data_train_scaled)) + 1) / 2)
selected_variables <- names(coef(fitf_bic))[!coef(fitf_bic) == 0]
ForwardVariables <- selected_variables[-1]
formula_2 <- as.formula(paste("Log_Salary ~", paste(ForwardVariables, collapse = " + ")))

#VALIDATION
#AIC
Forward_model_AIC <- lm(formula_1, data = data_valid_scaled)
#BIC
Forward_model_BIC <- lm(formula_2, data = data_valid_scaled)
Forward_model_AIC_predictions <- predict(Forward_model_AIC, newdata = data_valid_scaled)
Forward_model_AIC_predictions_mse <- mean((Forward_model_AIC_predictions - data_valid_scaled$Log_Salary) ^ 2)
Forward_model_BIC_predictions <- predict(Forward_model_BIC, newdata = data_valid_scaled)
Forward_model_BIC_predictions_mse <- mean((Forward_model_BIC_predictions - data_valid_scaled$Log_Salary) ^ 2)

  # Store the best model in the results data frame
  if (Forward_model_AIC_predictions_mse < Forward_model_BIC_predictions_mse) {
    best_model<- "AIC"
    FORWARD_MODEL <- lm(formula_1, data = data_test_scaled)
   aic_count_forward <- aic_count_forward + 1

  } else {
    best_model <- "BIC"
    FORWARD_MODEL <- lm(formula_2, data = data_test_scaled)
    bic_count_forward <- bic_count_forward + 1
  }


#TEST
forward_model_predictions <- predict(FORWARD_MODEL, newdata = data_test_scaled)
forward_model_predictions_mse <- mean((forward_model_predictions - data_test_scaled$Log_Salary)^2)
FORWARD_test_MSE_mean <- FORWARD_test_MSE_mean + forward_model_predictions_mse
}

MEAN_MSE_FORWARD <- FORWARD_test_MSE_mean / num_iterations
table_FORWARD <- data.frame(Model = c("AIC", "BIC"), Count = c(aic_count_forward, bic_count_forward))
#print(table_FORWARD)

```




```{r part15_b,results="hide"}
### 4- Step-wise approaches (and/or Recursive Feature Elimination).
##BACKWARD
Backward_test_MSE_mean<- 0
aic_count_backward <- 0
bic_count_backward <- 0

for (ii in 1:num_iterations) {
#Split 
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)

#TRAINING
#AIC
fit00 = lm(Log_Salary ~. , data = data_train_scaled)
fitf = stepAIC(fit0, 
               direction = "backward", data = data_train_scaled, k = log(nrow(data_train_scaled)))
selected_variables <- names(coef(fitf))[!coef(fitf) == 0]
#BIC 
fit00 = lm(Log_Salary ~ ., data = data_train_scaled)
fitf_bic = stepAIC(fit00, 
               direction = "backward", criterion = "BIC", data = data_train_scaled, k = log(nrow(data_train_scaled)))

selected_variables <- names(coef(fitf_bic))[!coef(fitf_bic) == 0]
#VALIDATION
#AIC
Backward_model_AIC <- lm(formula_1, data = data_valid_scaled)
Backward_model_BIC <- lm(formula_2, data = data_valid_scaled)
Backward_model_AIC_predictions <- predict(Backward_model_AIC, newdata = data_valid_scaled)
Backward_model_AIC_predictions_mse <- mean((Backward_model_AIC_predictions - data_valid_scaled$Log_Salary) ^ 2)
Backward_model_BIC_predictions <- predict(Backward_model_BIC, newdata = data_valid_scaled)
Backward_model_BIC_predictions_mse <- mean((Backward_model_BIC_predictions - data_valid_scaled$Log_Salary) ^ 2)
  # Store the best model in the results data frame
  if (Backward_model_AIC_predictions_mse < Backward_model_BIC_predictions_mse) {
    best_model<- "AIC"
    Backward_MODEL <- lm(formula_1, data = data_test_scaled)
   aic_count_backward <- aic_count_backward + 1

  } else {
    best_model <- "BIC"
    Backward_MODEL <- lm(formula_2, data = data_test_scaled)
    bic_count_backward <- bic_count_backward + 1
  }
#Test
Backward_model_predictions <- predict(Backward_MODEL, newdata = data_test_scaled)
Backward_model_predictions_mse <- mean((Backward_model_predictions - data_test_scaled$Log_Salary)^2)
Backward_test_MSE_mean <- Backward_test_MSE_mean + Backward_model_predictions_mse
}
# Calculate the average test MSE across all iterations for Least Squares
MEAN_MSE_BACKWARD <- Backward_test_MSE_mean / num_iterations
table_BACK <- data.frame(Model = c("AIC", "BIC"), Count = c(aic_count_backward, bic_count_backward))

#print(table_BACK)
```


```{r part16}
### 5- Lasso.
LASSO_test_MSE_mean <- 0
LASSO_BEST_alpha_values <- numeric(num_iterations)
LASSO_BEST_mse_values <- numeric(num_iterations)

for (i in 1:num_iterations) {
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)
  
#TRAIN
LASSO_alpha_values <- seq(0.1, 1, by = 0.1) 
LASSO_model_list <- list()  
LASSO_mse_values <- 0 

for (alpha in LASSO_alpha_values) {
  LASSO_model <- cv.glmnet(as.matrix(data_train_scaled), y_train_centered, alpha = alpha)
  LASSO_model_list[[as.character(alpha)]] <- LASSO_model
}

#VALIDATE
for (ii in seq_along(LASSO_alpha_values)) {
  LASSO_valid_predictions <- predict(LASSO_model_list[[as.character(LASSO_alpha_values[ii])]], newx = as.matrix(data_valid_scaled))
  LASSO_mse_values[ii] <- mean((LASSO_valid_predictions - y_valid_centered)^2)
}
LASSO_best_alpha <- LASSO_alpha_values[which.min(LASSO_mse_values)]
LASSO_best_mse<- min(LASSO_mse_values)
LASSO_best_model <- LASSO_model_list[[as.character(LASSO_best_alpha)]]
LASSO_BEST_alpha_values[i] <- LASSO_best_alpha
LASSO_BEST_mse_values[i] <- LASSO_best_mse

#TEST
LASSO_test_predictions <- predict(LASSO_best_model, newx = as.matrix(data_test_scaled))
LASSO_test_mse <- mean((LASSO_test_predictions - y_test_centered)^2)
LASSO_test_MSE_mean <- LASSO_test_MSE_mean + LASSO_test_mse

coefs03 = cbind(fit0$coef, fitl_03$beta)
colnames(coefs03) = c("Without L1", "With L1")
rownames(as.matrix(round(coefs03, 10)))[as.matrix(round(coefs03, 10))[, 2] != 0]
}

MEAN_MSE_LASSO <- LASSO_test_MSE_mean / num_iterations
table_LASSO <- data.frame(Iteration = 1:num_iterations, Best_Alpha = LASSO_BEST_alpha_values, Best_MSE = LASSO_BEST_mse_values)
#print(table_LASSO)

```


```{r part17}
### 6- Elastic Net.
results_ELASTIC <- data.frame(
  Iteration = numeric(num_iterations),
  Best_Alpha = numeric(num_iterations),
  Best_Lambda = numeric(num_iterations),
  Best_MSE = numeric(num_iterations),
  Test_Error = numeric(num_iterations))

for (i in 1:num_iterations) {
train_indices <- sample(1:N, 0.6 * N)  # 60% for training
remaining_indices <- setdiff(1:N, train_indices)
valid_indices <- sample(remaining_indices, 0.5 * length(remaining_indices))  # 20% for validation
test_indices <- setdiff(remaining_indices, valid_indices)  # 20% for test
# Center and scale the data.  the target variable
y_mean <- mean(as.numeric(organized_data[train_indices, ][, 1]))
y_train_centered <- as.numeric(organized_data[train_indices, ][, 1]) - y_mean
y_valid_centered <- as.numeric(organized_data[valid_indices, ][, 1]) - y_mean
y_test_centered <- as.numeric(organized_data[test_indices, ][, 1]) - y_mean
# Center and standardize
X_train_scaled <- as.data.frame(scale(as.matrix(organized_data[train_indices, ][, -1]), center = TRUE, scale = TRUE))
X_valid_scaled <- as.data.frame(scale(as.matrix(organized_data[valid_indices, ][, -1]), center = TRUE, scale = TRUE))
X_test_scaled <- as.data.frame(scale(as.matrix(organized_data[test_indices, ][, -1]), center = TRUE, scale = TRUE))
data_train_scaled <- cbind(Log_Salary = y_train_centered, X_train_scaled)
data_valid_scaled <- cbind(Log_Salary = y_valid_centered, X_valid_scaled)
data_test_scaled <- cbind(Log_Salary = y_test_centered, X_test_scaled)
  

#TRAIN
# Define a range of alpha values (from 0.1 to 1)
alpha_values <- seq(0.1, 1, by = 0.1)
  best_model <- NULL
  best_alpha <- NULL
  best_lambda <- NULL
  best_validation_error <- Inf

for (alpha in alpha_values) {
  model <- glmnet(as.matrix(X_train_scaled), y_train_centered, alpha = alpha)
  cv_model <- cv.glmnet(as.matrix(X_train_scaled), y_train_centered, alpha = alpha)
  best_lambda <- cv_model$lambda.min
  predictions_val <- predict(model, newx = as.matrix(X_valid_scaled), s = best_lambda)
  validation_error <- mean((predictions_val - y_valid_centered)^2)
  if (validation_error < best_validation_error) {
    best_validation_error <- validation_error
    best_model <- model  
    best_model <- model 
    best_alpha <- alpha  
  }
}

#TEST
best_lambda <- cv.glmnet(as.matrix(X_train_scaled), y_train_centered, alpha = alpha)$lambda.min
predictions_test <- predict(best_model, newx = as.matrix(X_test_scaled), s = best_lambda)
test_error <- mean((predictions_test - y_test_centered)^2)
  results_ELASTIC[i, "Iteration"] <- i
  results_ELASTIC[i, "Best_Alpha"] <- best_alpha
  results_ELASTIC[i, "Best_Lambda"] <- best_lambda
  results_ELASTIC[i, "Best_MSE"] <- best_validation_error
  results_ELASTIC[i, "Test_Error"] <- test_error
}
MEAN_MSE_ELASTIC <- mean(results_ELASTIC$Test_Error)
#print(results_ELASTIC)

```

### (ii) Visualize the results of your comparisons. 

You can choose to only show the results with the best tuning parameter values.


```{r part18}

method_names <- c("Least Squares", "Ridge", "Best Subsets", "Forward","Backward", "Lasso", "Elastic Net")
mse_values <- c(MEAN_MSE_LS, MEAN_MSE_RIDGE, MEAN_MSE_BESTSUBSET, 
                MEAN_MSE_FORWARD, MEAN_MSE_BACKWARD, MEAN_MSE_LASSO, MEAN_MSE_ELASTIC)
results_df <- data.frame(Method = method_names, Average_Test_MSE = mse_values)
results_df$Method <- factor(results_df$Method, levels = method_names)
results_df <- results_df[order(results_df$Method), ]
ggplot(results_df, aes(x = Method, y = Average_Test_MSE)) +
  geom_bar(stat = "identity") +
  labs(title = "Comparison of Average Test MSE for Linear Methods",
       x = "Method", y = "Average Test MSE") +
  theme_minimal()

print(results_df)
```

### (iii) Reflection. 

Which types of methods give the best prediction error? Why do these methods
perform well? Do any methods seem to overfit to the training set? If so, why? Do all the
methods that give similar predictions choose the same subset of variables? Which is the
overall best method for prediction on this dataset? Explain and expand on your responses.

```{r part19_0}

## ANSWER:


cat("Given the mean and minimun values of the scaled and log normalized data,","\n",
"Least Squares method might be over fitting.","\n",
"for selected parameters, the order of MSE values from largest to smallest are; Best Subset,","\n",
"Ridge, Elastic net, Forward, Bacward, Elastic net and Lasso.  ","\n",
"given the approximate mean, max, min","\n",
"Lasso with about  looks the best fit for this results.", "\n",
"But depending on the question or the goal of the question,", "\n", 
"Forward and backward stepwise, and Elastic net can also be applicable." , "\n", 
"Mean of scaled data=",mean(DATA_SCALED$Log_Salary),"\n",
"Max of scaled data=",max(DATA_SCALED$Log_Salary),"\n",
"Mix of scaled data=",min(DATA_SCALED$Log_Salary),"\n", 
"Please see below tables to show details about the best hyperparameters or categories:")


print(table_RIDGE)
print(table_BESTSUBSET)
print(table_FORWARD)
print(table_BACK)
print(table_LASSO)
print(results_ELASTIC)

```

\newpage


### 2. Empirical Demonstrations of Regression Properties. 

For these problems, you can use the data mentioned above or you may need to design your own linear regression simulation.

### (a) Empirically demonstrate that fitting linear regression with an intercept term is equivalent to (i)fitting linear regression when centering Y and centering the columns of X, and (ii) fitting linear regression when adding a column of ones to X.

```{r part19}
y = as.numeric(organized_data1[, 1])
X = as.matrix(organized_data1[, -1])
y_centered = y - mean(y)
X_centered = scale(X, center = TRUE, scale = FALSE)
X_standardized = scale(X, center = TRUE, scale = TRUE)

model_centered <- lm(y_centered ~ X_centered , data = organized_data1)

X_with_intercept <- cbind(1, organized_data1[,-1])
model_with_ones <- lm(Salary ~ ., data = X_with_intercept)

summary(model_centered)
summary(model_with_ones)
```




### (b) Empirically demonstrate that the least squares solution has zero training error when p > n.

```{r part20}
set.seed(123)
n <- 50 
p <- 60  
X <- matrix(rnorm(n * p), ncol = p)
y <- rnorm(n)
lm_model <- lm(y ~ X)
train_error <- sum(lm_model$residuals^2)
# Check if the training error is approximately zero
if (train_error < 1e-10) {
  cat("The least squares solution has zero training error, when we use the cut off as train_error < 1e-10.")
} else {
  cat("The least squares solution does not have zero training error.")
}

```

### (c) Empirically demonstrate the MSE Existence Theorem. 

```{r part21}
set.seed(123)
n <- dim(organized_data1)[1]
train_index <- sample(1:n, 0.8 * n)
train_data <- organized_data1[train_index, ]
test_data <- organized_data1[-train_index, ]
# linear regression model
ols_model <- lm(Salary ~ ., data = train_data)
ols_predictions <- predict(ols_model, newdata = test_data)
mse_ols <- mean((test_data$Salary - ols_predictions)^2)
# Ridge regression
lambda_seq <- 10^seq(10, -2, length = 100)
ridge_models <- glmnet(as.matrix(train_data[, -1]), train_data$Salary, alpha = 0, lambda = lambda_seq)
mse_ridge <- sapply(lambda_seq, function(lambda) {
  ridge_predictions <- predict(ridge_models, s = lambda, newx = as.matrix(test_data[, -1]))
  mean((test_data$Salary - ridge_predictions)^2)
})
# Find the lambda with lower MSE than OLS
lambda_min_mse <- lambda_seq[which.min(mse_ridge)]
cat("MSE for Ridge=", min(mse_ridge), "\n")
cat("MSE for OLS:", mse_ols, "\n")
if (min(mse_ridge) < mse_ols) {
  cat("The MSE Existence Theorem is empirically demonstrated that there exists a value of LAMDA for which MSE Ridge is less than MSE LS.")
} else {
  cat("The MSE Existence Theorem is not empirically demonstrated.")
}

```




